---
title: "01_Create_Fragmentation_Profiles"
author: "Alice McAtamney"
date: "2023-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

* __Introduction:__
  + Quick paragraph on logic behind this script

* __Setup:__
  + Loads packages
  + Prepares sample info
  + Loads bam files


* __Generate Fragment Profiles: __ 
  + Function to calculate fragment counts
  + Loop to produce fragment profile for each sample

* __Plot Fragment Profiles: __ 
  + Plots for individual samples
  + Plots for fragment profiles combined by group


## Introduction

This script is an adaptation of the DELFI method created by Cristiano et. al (2019). It creates "fragment profiles" for each sample.
The functions in this script break the genome of each sample into 5 Mbp bins and calculate the number of short (100-150 bp) and long (151-220 bp) fragments in each bin, plotting this data for each individual sample, as well as samples grouped by their cancer status.

To run this script you will need:

* a csv file containing:
  + sample_id
  + bam_path (i.e. path/to/bam)
  + group (eg. high_crc, low_crc, healthy)
* output folder path

## Setup

```{r Setup}
library(GenomicAlignments)
library(BSgenome.Hsapiens.UCSC.hg19)
library(ggplot2)
library(tidyverse)
library(GenomicRanges)
```


```{r enter sample info}
#sample_info <- "path/to/sample_data.csv"
sample_info <- "~/genomic_features/data/bams/all_sample_info.csv"
#output <- "path/to/output"
output <-  "~/genomic_features/output/fragment_profiles"
#path to regions of genome to filter out eg. centromeres, low map regions
filter_path <- "/Volumes/archive/userdata/student_users/alicemcatamney/genomic_features/data/filters.hg19.rda"
```



```{r load sample info}
sample_info <- read_csv(sample_info)
samples <- sample_info$sample_id
bams <- sample_info$bam_path

head(samples)
```



```{bash create folders to organise output, eval = FALSE}
# change to whatever your output directory is
# cd $output
mkdir granges
mkdir counts
mkdir graphs
```


It takes a long time to read in bams as GRange, so reading in and saving as RDS file for later use.
Set eval = TRUE if this has not been done for your samples

```{r load bams as GRange function, eval = FALSE}
load_bam_as_GRange <- function(sample, bam_path, save = TRUE, save_path){
  print(paste("Loading", sample, "as GRange."))
  param <- ScanBamParam(flag = scanBamFlag(isDuplicate = FALSE,
                                           isSecondaryAlignment = FALSE,
                                           isUnmappedQuery = FALSE),
                        mapqFilter = 30)
  
  gp <- readGAlignmentPairs(bam_path, param = param)
  gr <- granges(keepSeqlevels(gp, paste0("chr", 1:22), pruning.mode="coarse"),
                on.discordant.seqnames="drop")
  if(save == TRUE){
    print("Saving GRange")
    saveRDS(gr, save_path)
  }
}


for(i in 1:length(samples)){
  save_path <- paste0(output, "/granges/", samples[i], ".gr.RDS")
  load_bam_as_GRange(samples[i], bam_path = bams[i], save_path = save_path)
}

```


## Generate Fragment Profiles

This function counts number of short and long frags in 5Mbp windows getting fragment profiles data.
This can also take a while, so again only set eval = TRUE if this has not been done for your samples.

If plot density == TRUE, the function will also produce a density plot that shows the distribution of your fragment lengths.

```{r function get fragment counts function, eval = FALSE}

fragment_counts <- function(sample_id, gr, output, filters, plot_density = TRUE){
  w.all <- width(gr)
  q.all <- quantile(w.all, c(0.001, 0.999))
  gr <- gr[which(w.all > q.all[1] & w.all < q.all[2])]
  w.all <- width(gr)
  gr.clean <- gr[-queryHits(findOverlaps(gr, filters.hg19))]
  lengths <- data.frame(length = w.all)
  
  if(plot_density == TRUE){
    uw <- unique(w.all)
    modal_width <- uw[which.max(tabulate(match(w.all, uw)))]
    dp <- ggplot(lengths, aes(x = length)) +
      geom_density() +
      geom_vline(xintercept = modal_width, linewidth = 0.2)
    ggsave(filename = paste0(output, "/graphs/", sample_id, ".dp.png"), plot = dp, device = "png")
  }
  
  print("Filtering")
  w.all <- width(gr.clean)
  fragments <- gr.clean[which(w.all >= 100 & w.all <= 220)]
  w <- width(fragments)
  bins <- tileGenome(seqinfo(Hsapiens),
                     tilewidth = 5000000,
                     cut.last.tile.in.chrom = TRUE)
  chrms <- paste0("chr", 1:22)
  bins <- bins[seqnames(bins) %in% chrms]
  
  print("Counting Fragments in each bin")
  frag.list <- split(fragments, w)
  counts <- sapply(frag.list, function(x) countOverlaps(bins, x)) 
  #put counts into matrix
  if(min(w) > 100) {
    m0 <- matrix(0, ncol=min(w) - 100, nrow=nrow(counts),
                 dimnames=list(rownames(counts), 100:(min(w)-1)))
    counts <- cbind(m0, counts)
  }
  
  short <- rowSums(counts[,1:51]) #adds up number of fragments between 100-150bp for each row/bin
  long <- rowSums(counts[,52:121])
  bins$short <- short
  bins$long <- long
  bins$ratio <- short/long
  bins$nfrags <- short+long
  
  saveRDS(data.frame(bins), paste0(output, "/counts/", sample_id, ".counts.RDS"))
}

```

This section runs the function above
```{r run fragment counts, eval = FALSE}
# filters.hg19 obtained from Delfi script 00-filtered_regions.r
filters <- load(filter_path)

for(i in 1:length(samples)){
  print(samples[i])
  gr <- readRDS(paste0(output, "/granges/", samples[i], ".gr.RDS"))
  fragment_counts(sample_id = samples[i], gr = gr, output = output, filters = filters, plot_density = TRUE)
}


```



## Plot fragment profiles

```{r plot individual frag profiles, eval = FALSE}

for(i in samples){
  bin.df <- readRDS(paste0(output, "/counts/", i, ".counts.RDS"))
  fp <- ggplot(bin.df, aes(start, ratio)) +
        geom_line(linewidth = 0.2) +
        labs(y = "Ratio") +
        facet_grid(.~seqnames,
                   scales = "free",
                   space = "free",
                   switch = "both") +
        scale_x_continuous(breaks = round(seq(min(bin.df$start),
                                              max(bin.df$start),
                                              by = 100000000),1)) +
        theme_minimal() +
        theme(axis.text.x = element_blank(),
              axis.ticks.x = element_blank(),
              axis.title.x = element_blank(),
              axis.line = element_line(color = "black"),
              strip.background = element_rect(
                color = "white", fill = "white"),
              strip.placement = "outside",
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank())
  
ggsave(plot = fp,
       filename = paste0(output, "/graphs/", i, ".fp.png"),
       device = "png",
       width = 12, height = 1.5, units = "in")
  
}


```




The function below can be used to plot fragment profiles separated by group. Function requires:

* sample_ids:
  * A character vector of sample ids with fps you want to combine.
  * Samples must have corresponding count data in "output/counts" named [id].counts.RDS
* mean.colour:
  * string indicating the colour you want the mean line to be
* output:
  * path to output directory.
  * directory must contain subdirectory named counts with corresponding fp counts data (generated above)
  * directory must contain subdirectory with subdirectory called graphs
* group.name
  * a string indicating the category that the fps you are combining fall into
  * eg. if you are combining healthy control fps, enter "controls" and final plot will be called "controls.scaled.fp.png"
* scale:
  * this is automatically set to false. This option indicates whether you want the graphs to be zero-centered, or have their true median.


```{r function combine frag profiles}

combine_fps <- function(sample.ids, mean.colour, output, group.name, scale = FALSE){
  
  # combine data into one tibble
  paths <- paste0(output, "/counts/", sample.ids, ".counts.RDS")
  bins.list <- lapply(paths, readRDS)
  tib.list <- lapply(bins.list, as_tibble)
  names(tib.list) <- sample.ids
  
  if(scale == TRUE){
    for(i in 1:length(tib.list)){
      tib.list[[i]] <- mutate(tib.list[[i]], ratio = scale(ratio)[,1])
    }
  }
  combined.counts <- map2(tib.list, names(tib.list), ~ mutate(.x, id = .y)) %>%
    bind_rows() %>% dplyr::select(id, everything())
  
  if(scale == FALSE){
    saveRDS(combined.counts, paste0(output, "/counts/", group.name, ".df.RDS"))
  } else {
    saveRDS(combined.counts, paste0(output, "/counts/", group.name, ".df.scaled.RDS"))
  }
  
  # get average
  bin.averages <- combined.counts %>%
    group_by(start, seqnames) %>%
    summarise(mean.ratio = mean(ratio))
  
  # fragment profile ggplot theme
  theme_fp <- theme_minimal() +
              theme(axis.text.x = element_blank(),
                    axis.ticks.x = element_blank(),
                    axis.title.x = element_blank(),
                    axis.line = element_line(color = "black"),
                    strip.background = element_rect(color = "white", fill = "white"),
                    strip.placement = "outside",
                    #strip.text.x.bottom = element_text(angle = 90),
                    panel.grid.major = element_blank(),
                    panel.grid.minor = element_blank())
  
  combined.fp.plot <- ggplot(combined.counts, aes(start, ratio)) +
      geom_line(aes(group = id), color = "#737373", size = 0.2) +
      labs(y = "Ratio") +
      facet_grid(.~seqnames,
               scales = "free",
               space = "free",
               switch = "both") +
      scale_x_continuous(breaks = round(seq(min(combined.counts$start),
                                            max(combined.counts$start),
                                            by = 100000000),1)) +
      theme_fp +
      ylim(0,0.8)
  
  combined.fp.plot + geom_line(data = bin.averages,
                             aes(start, mean.ratio),
                             size = 0.5, color = mean.colour)
  
  if(scale == FALSE){
    ggsave(filename = paste0(output, "/graphs/", group.name, ".unscaled.fp.png"),
           device = "png",
           width = 12,
           height = 1.5,
           units = "in")
  } else {
    ggsave(filename = paste0(output, "/graphs/", group.name, ".scaled.fp.png"),
           device = "png",
           width = 12,
           height = 1.5,
           units = "in")
  }

}



```





```{r combine frag profiles, eval = FALSE}
combine_fps(sample.ids = samples,
            mean.colour = "grey",
            output = output,
            group.name = "all",
            scale = FALSE)

sample_info$group

## BRCA
brca_info <- sample_info[sample_info$group == "my_brca",]
brca_samples <- brca_info$sample_id
combine_fps(sample.ids = brca_samples,
            mean.colour = "pink",
            output = output,
            group.name = "brca")
combine_fps(sample.ids = brca_samples,
            mean.colour = "pink",
            output = output,
            group.name = "brca",
            scale = TRUE)

## my CRC
# high
my_crc_info <- sample_info[sample_info$group == "my_crc",]
my_crc_info <- my_crc_info %>%
  mutate(group = ifelse(substr(sample_id, nchar(sample_id), nchar(sample_id)) == "H",
                         "high", "low"))
my_crc_high <- my_crc_info[my_crc_info$group == "high",]
combine_fps(sample.ids = my_crc_high$sample_id,
            mean.colour = "red",
            output = output,
            group.name = "my_crc_high")
combine_fps(sample.ids = my_crc_high$sample_id,
            mean.colour = "red",
            output = output,
            group.name = "my_crc_high",
            scale = TRUE)

# low
my_crc_low <- my_crc_info[my_crc_info$group == "low",]
combine_fps(sample.ids = my_crc_low$sample_id,
            mean.colour = "blue",
            output = output,
            group.name = "my_crc_low")
combine_fps(sample.ids = my_crc_low$sample_id,
            mean.colour = "blue",
            output = output,
            group.name = "my_crc_low",
            scale = TRUE)


# Delfi CRC
delfi_crc_info <- sample_info[sample_info$group == "delfi_crc",]

combine_fps(sample.ids = delfi_crc_info$sample_id,
            mean.colour = "black",
            output = output,
            group.name = "delfi_crc",
            scale = FALSE)
combine_fps(sample.ids = delfi_crc_info$sample_id,
            mean.colour = "orange",
            output = output,
            group.name = "delfi_crc",
            scale = TRUE)

# Healthy
delfi_healthy_info <- sample_info[sample_info$group == "healthy",]

combine_fps(sample.ids = delfi_healthy_info$sample_id,
            mean.colour = "black",
            output = output,
            group.name = "delfi_healthy",
            scale = FALSE)
combine_fps(sample.ids = delfi_healthy_info$sample_id,
            mean.colour = "black",
            output = output,
            group.name = "delfi_healthy",
            scale = TRUE)


```
